---
title: "A brief manual"
author: Joshua N. Pritikin
output:
  html_document:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{A brief manual}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
library(ggplot2)
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
set.seed(1)
```

# Overview

The **pcFactorStan** package for **R** provides convenience functions and pre-programmed **Stan** models related to analysis of pairwise comparison data. Its purpose is to make fitting models using Stan easy. **pcFactorStan** relies on the **rstan** package, which should be installed first.
[See here](https://github.com/stan-dev/rstan/wiki/RStan-Getting-Started) for instructions on installing **rstan**.

One situation where a factor model might be useful
is when there are people that play in tournaments of more than
one board game. For example, the computer player AlphaZero (Silver
et al. 2018) has trained to play chess, shogi, and Go. We can take
the tournament match outcome data and find rankings among the
players for each of these games. We may also suspect that there is
a latent board game skill that accounts for some proportion of the
variance in the per-board game rankings. This proportion can be
recovered by the factor model.

Our goal may be to fit a factor model, but it is necessary to
build up the model step-by-step.
There are essentially three models: 'unidim', 'covariance', and 'factor'.
'unidim' analyzes a single item.
'covariance' is suitable for two or more items.
Once you have vetted your items with the 'unidim' and 'covariance'
models, then you can try the 'factor' model.
There is also a special model 'unidim+adapt'.
Except for this model, the other models require a scaling constant.
To find an appropriate scaling constant, we will
fit 'unidim+adapt' to each item separately
and then take the median of median point estimates
to set the scale.

# Brief tutorial

## Physical activity flow propensity

The R code below first loads **pcFactorStan**, which implicitly loads **rstan**. The two commands that follow set generally recommended options related to **rstan**. The first causes compiled **Stan** models to be cached on disk, allowing models to run more quickly after the first time. The second causes **Stan** to run multiple chains in parallel.

```{r, message=FALSE}
library(pcFactorStan)
rstan_options(auto_write = TRUE)
options(mc.cores = parallel::detectCores())
```

Next we take a peek at the data.

```{r, results='asis'}
kable(head(phyActFlowPropensity))
```

These data consist of paired comparisons of 87 physical activities on 16 flow-related facets. Participants submitted two activities using free-form input. These activities were substitute into item templates. For example, the 'predict' item asked, "How predictable is the action?" with response options:

* A1 is much more predictable than A2.
* A1 is somewhat more predictable than A2.
* Both offer roughly equal predictability.
* A2 is somewhat more predictable than A1.
* A2 is much more predictable than A1.

A "somewhat more" response is denoted with a 1 or -1
and "much more" with a 2 or -2.
A tie (i.e. "roughly equal") is denoted by a zero.
We will need to analyze each item separately before
we analyze them together. Therefore, we will start
with 'skill'.

Data is fed into **Stan** in a partially digested form, and the next block of code demonstrates how a suitable data object may be constructed using the `prepData()` function. This function automatically determines how
many threshold parameters to estimate based on the
range of your data.
One thing it does not do is pick a `varCorrection` factor. The `varCorrection` determines the degree of adaption in the model. Usually setting 2.0 or 3.0 will obtain optimal results.

```{r}
dl <- prepData(phyActFlowPropensity[,c(paste0('pa',1:2), 'skill')])
dl$varCorrection <- 2.0
```

Next we fit the model using the `pcStan()` function, which is a wrapper for `stan()` from **rstan**. We also choose the number of chains. By default, the first half of each chain is discarded as warm up.

```{r, message=FALSE, results='hide'}
fit1 <- pcStan(data=dl, chains=4L)
```

A variety of diagnostics are available to check whether the sampler ran into trouble.

```{r}
check_hmc_diagnostics(fit1)
```

Everything looks good, but there are a few more things to check.
We want $\widehat R < 1.015$ and effective sample size greater than 100 times the number of chains (Vehtari et al., 2019).

```{r}
allPars <- summary(fit1, probs=c())$summary
print(min(allPars[,'n_eff']))
print(max(allPars[,'Rhat']))
```

Again, everything looks good. If the targets values were not reached
then we would sample the model again with more iterations.
Time for a plot,

```{r}
theta <- summary(fit1, pars=c("theta"), probs=c())$summary[,'mean']
palist <- levels(phyActFlowPropensity$pa1)

ggplot(data.frame(x=theta, activity=palist, y=0.47)) +
  geom_point(aes(x=x),y=0) +
  geom_text(aes(label=activity, x=x, y=y),
            angle=85, hjust=0, size=2,
            position = position_jitter(width = 0, height = 0.4)) + ylim(0,1) +
  theme(legend.position="none",
        axis.title.x=element_blank(),
        axis.title.y=element_blank(),
        axis.text.y=element_blank(),
        axis.ticks.y=element_blank())
```

That seems like a fairly reasonable ranking for skill.
As pretty as the plot is, the main reason that we fit this model
was to find a scaling factor that would produce a standard
deviation close to 1.0,

```{r}
s50 <- summary(fit1, pars=c("scale"), probs=c(.5))$summary[,'50%']
print(s50)
```

We use the median instead of the mean because `scale` is not likely to have a symmetric marginal posterior distribution.

```{r, message=FALSE, results='hide'}
result <- calibrateItems(phyActFlowPropensity)

```

```{r, results='asis'}
kable(result)
```


# Technical notes

Given that my background is more in programming than math,
I am not a fan of the greek letters used by
mathematicians with such enthusiasm.
When I name variables, I favor the expressive over the succinct.
If you read through the **Stan** models, you will find some
variables prefixed with `raw`. These are special variables
internal to the model. In particular, you should not try
to evaluate the $\widehat R$ or effective sample size
of `raw` parameters.

## Unidim

| parameter | prior        | purpose                  |
|-----------|--------------|--------------------------|
| threshold | normal(0,2)  | item response thresholds |
| theta     | normal(0,1)  | latent score    |

The 'unidim+adapt' model has a `varCorrection` constant
that is used to calibrate the `scale`. For multivariate models,
`scale` should be set to the median of the item-wise scales.

## Covariance

| parameter | prior          | purpose                              |
|-----------|----------------|--------------------------------------|
| threshold | normal(0,2)    | item response thresholds             |
| thetaCor  | lkj(2)         | correlations between items           |
| sigma     | lognormal(1,1) | relative item standard deviations    |
| theta     | _see below_    | latent score                |

Thresholds for all
items are combined into a single vector.
The prior for `theta` is multivariate normal with correlations
`thetaCor` and standard deviations `sigma`.

## Factor

| parameter      | prior               | purpose                              |
|----------------|---------------------|--------------------------------------|
| threshold      | normal(0,2)         | item response thresholds             |
| unique         | normal(1,1) T[0,]   | standard deviation of unique scores  |
| uniqueTheta    | normal(0,1)         | unique scores                        |
| factorLoadings | normal(0,1)         | signed standard deviation of factor scores  |
| factor         | normal(0,1)         | factor scores                        |
| factorProp     | N/A                 | signed factor variance proportion    |

Thresholds for all items are combined into a single vector.
`factorProp` is computed using Equation 3 of Gelman et al. (in press)
and has no prior of its own.
`factorLoadings` is in standard deviation units but can be negative.
Similarly, `factorProp` is a signed proportion bounded between -1 and 1.

# Writing your own **Stan** models

We anticipate that **pcFactorModel** users may eventually want to write their own **Stan** models. The models included in **pcFactorStan** may be used as a starting point in developing custom models. The following code shows how to find and view one of these files.

```{r, results='hide'}
cat(readLines(locateModel('unidim')), sep = "\n")
```

# References

Gelman, A., Goodrich, B., Gabry, J., & Vehtari, A. (in press). R-squared for Bayesian regression models. The American Statistician. DOI: 10.1080/00031305.2018.1549100

Silver, D., Hubert, T., Schrittwieser, J., Antonoglou, I., Lai, M.,
Guez, A., ... & Lillicrap, T. (2018). A general reinforcement
learning algorithm that masters chess, shogi, and Go through
self-play. Science, 362(6419), 1140-1144.

Vehtari, A., Gelman, A., Simpson, D., Carpenter, B., & Bürkner, P. C. (2019). Rank-normalization, folding, and localization: An improved $\widehat R$ for assessing convergence of MCMC. arXiv preprint arXiv:1903.08008.
