---
title: "A brief manual"
author: Joshua N. Pritikin
output:
  html_document:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{Getting started with pcFactorStan}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
set.seed(1)
is_CRAN <- !identical(Sys.getenv("NOT_CRAN"), "true")
if (!is_CRAN) {
   options(mc.cores = parallel::detectCores())
} else {
   print("CRAN detected")
   q()  # takes too long, pretend everything is fine
}
library(knitr)
library(ggplot2)
library(reshape2)
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  cache.lazy = FALSE  # https://github.com/yihui/knitr/issues/572
)
options(digits=4)
options(scipen=2)
```

# Overview

The **pcFactorStan** package for **R** provides convenience functions and pre-programmed **Stan** models related to analysis of paired comparison data. Its purpose is to make fitting models using Stan easy and easy to understand. **pcFactorStan** relies on the **rstan** package, which should be installed first.
[See here](https://github.com/stan-dev/rstan/wiki/RStan-Getting-Started) for instructions on installing **rstan**.

One situation where a factor model might be useful
is when there are people that play in tournaments of more than
one game. For example, the computer player AlphaZero (Silver
et al. 2018) has trained to play chess, shogi, and Go. We can take
the tournament match outcome data for each of these games
and find rankings among the players. We may also suspect that there is
a latent board game skill that accounts for some proportion of the
variance in the per-board game rankings. This proportion can be
recovered by the factor model.

Our goal may be to fit a factor model, but it is necessary to
build up the model step-by-step.
There are essentially three models: 'unidim', 'correlation', and 'factor'.
'unidim' analyzes a single item.
'correlation' is suitable for two or more items.
Once you have vetted your items with the 'unidim' and 'correlation'
models, then you can try the 'factor' model.
There is also a special model 'unidim_adapt'.
Except for this model, the other models require scaling constants.
To find appropriate scaling constants, we will
fit 'unidim_adapt' to each item separately.

# Brief tutorial

## Physical activity flow propensity

The R code below first loads **rstan** and **pcFactorStan**.
For extra diagnostics, we also load **loo**.

```{r, message=FALSE}
library(rstan)
library(pcFactorStan)
library(loo)
```

Next we take a peek at the data.

```{r, results='hide'}
head(phyActFlowPropensity)
```
```{r, results='asis', echo=FALSE}
kable(head(phyActFlowPropensity))
```

These data consist of paired comparisons of 87 physical activities on 16 flow-related facets. Participants submitted two activities using free-form input. These activities were substituted into item templates. For example, Item _predict_ consisted of the prompt, "How predictable is the action?" with response options:

* `A1` is much more predictable than `A2`.
* `A1` is somewhat more predictable than `A2`.
* Both offer roughly equal predictability.
* `A2` is somewhat more predictable than `A1`.
* `A2` is much more predictable than `A1`.

If the participant selected 'golf' and 'running' for activities then 'golf' was substituted into `A1` and 'running' into `A2`. Duly prepared, the item was presented and the participant asked to select the most plausible statement.

A _somewhat more_ response is scored 1 or -1
and _much more_ scored 2 or -2.
A tie (i.e. _roughly equal_) is scored as zero.
We will need to analyze each item separately before
we analyze them together. Therefore, we will start
with Item _skill_.

Data must be fed into **Stan** in a partially digested form. The next block of code demonstrates how a suitable data list may be constructed using the `prepData()` function. This function automatically determines the
number of threshold parameters based on the
range observed in your data.
One thing it does not do is pick a `varCorrection` factor. The `varCorrection` determines the degree of adaption in the model. Usually some choice between 2.0 to 4.0 will obtain optimal results.

```{r}
dl <- prepData(phyActFlowPropensity[,c(paste0('pa',1:2), 'skill')])
dl$varCorrection <- 2.0
```

Next we fit the model using the `pcStan()` function, which is a wrapper for `stan()` from **rstan**. We also choose the number of chains.
As is customary **Stan** procedure, the first half of each chain is used to estimate the
sampler's weight matrix (i.e. warm up) and excluded from inference.

```{r pcStan, message=FALSE, results='hide', cache=TRUE}
fit1 <- pcStan("unidim_adapt", data=dl)
```

A variety of diagnostics are available to check whether the sampler ran into trouble.

```{r pcStanDiag1, cache=TRUE}
check_hmc_diagnostics(fit1)
```

Everything looks good, but there are a few more things to check.
We want $\widehat R$ < 1.015 and effective sample size greater than 100 times the number of chains (Vehtari et al., 2019).

```{r pcStanDiag2, cache=TRUE}
allPars <- summary(fit1, probs=c())$summary 
print(min(allPars[,'n_eff']))
print(max(allPars[,'Rhat']))
```

Again, everything looks good. If the target values were not reached
then we would sample the model again with more iterations.
Time for a plot,

```{r skill, cache=TRUE}
library(ggplot2)

theta <- summary(fit1, pars=c("theta"), probs=c())$summary[,'mean']

ggplot(data.frame(x=theta, activity=dl$nameInfo$pa, y=0.47)) +
  geom_point(aes(x=x),y=0) +
  geom_text(aes(label=activity, x=x, y=y),
            angle=85, hjust=0, size=2,
            position = position_jitter(width = 0, height = 0.4)) + ylim(0,1) +
  theme(legend.position="none",
        axis.title.x=element_blank(),
        axis.title.y=element_blank(),
        axis.text.y=element_blank(),
        axis.ticks.y=element_blank())
```

Intuitively, this seems like a fairly reasonable ranking for skill.
As pretty as the plot is, the main reason that we fit this model
was to find a scaling constant to produce a standard
deviation close to 1.0,

```{r}
s50 <- summary(fit1, pars=c("scale"), probs=c(.5))$summary[,'50%']
print(s50)
```
```{r, results='hide', include = FALSE}
rm(fit1)  # free up some memory
```

We use the median instead of the mean because `scale` is not likely to have a symmetric marginal posterior distribution.
We obtained `r s50`, but that value is just for one item.
We have to perform the same procedure for every item.
Wow, that would be really tedious ... if we did not have a function to do it for us!
Fortunately, `calibrateItems` takes care of it and produces a table
of the pertinent data,

```{r calibrateItems, message=FALSE, results='hide', cache=TRUE}
result <- calibrateItems(phyActFlowPropensity, iter=1000L) 
```
```{r, results='hide'}
print(result)
```
```{r, results='asis', echo=FALSE}
kable(result)
```

Item _goal1_ ran into trouble.
A nonzero count of divergent transitions or low_bfmi means that the item
contained too little signal to estimate.
Item _feedback1_ is also prone to failure.
We could try again with `varCorrection=1.0`,
but we are going to exclude these items instead.
The model succeeded on the rest of the items.
I requested `iter=1000L` to demonstrate how `calibrateItems` will resample the model until the `n_eff` is large enough and the `Rhat` small enough.
As demonstrated in the _iter_ column, some items needed more than 1000 samples to converge.

Next we will fit the correlation model. We exclude the Cholesky
factor of the correlation matrix `rawThetaCorChol` because the
regular correlation matrix is also output.
```{r covarianceData, cache=TRUE}
pafp <- phyActFlowPropensity 
excl <- match(c('goal1','feedback1'), colnames(pafp))
pafp <- pafp[,-excl]
dl <- prepData(pafp)
dl$scale <- result[-excl,'scale'] 
```

```{r covariance, message=FALSE, results='hide', cache=TRUE}
fit2 <- pcStan("correlation", data=dl, include=FALSE, pars=c('rawTheta', 'rawThetaCorChol'))
```

```{r covarianceDiag1, cache=TRUE}
check_hmc_diagnostics(fit2)

allPars <- summary(fit2, probs=0.5)$summary 
print(min(allPars[,'n_eff']))
print(max(allPars[,'Rhat']))
```
The HMC diagnostics look good, but ... oh dear!
Something is wrong with the `n_eff` and $\widehat R$.
Let us look more carefully,

```{r covarianceDiag2, cache=TRUE}
head(allPars[order(allPars[,'sd']),])
```

Ah ha! It looks like all the entries of the correlation matrix are reported,
including the entries that are not stochastic but are fixed to constant values.
We need to filter those out to get sensible results.

```{r covarianceDiag3, cache=TRUE}
allPars <- allPars[allPars[,'sd'] > 1e-6,] 
print(min(allPars[,'n_eff']))
print(max(allPars[,'Rhat']))
```

Ah, much better. Now we can inspect the correlation matrix. There are
many ways to visualize a correlation matrix. One of my favorite
ways is to plot it using the **qgraph** package,

```{r covPlot, cache=TRUE}
covItemNames <- dl$nameInfo$item 
tc <- summary(fit2, pars=c("thetaCor"), probs=c(.5))$summary[,'50%']
tcor <- matrix(tc, length(covItemNames), length(covItemNames))
dimnames(tcor) <- list(covItemNames, covItemNames)

library(qgraph)
qgraph(tcor, layout = "spring", graph = "cor", labels=colnames(tcor),
       legend.cex = 0.3,
       cut = 0.3, maximum = 1, minimum = 0, esize = 20,
       vsize = 7, repulsion = 0.8, negDashed=TRUE, theme="colorblind")
```

Based on this plot and theoretical considerations,
I decided to exclude _spont_, _control_, _evaluated_, and _waiting_
from the factor model.
A detailed rationale for why these items, and not others, are excluded
will be presented in a forthcoming article.
For now, let us focus on the mechanics of data analysis.
Here are item response curves,

```{r responseCurves, cache=TRUE}
df <- responseCurve(dl, fit2,
  item=setdiff(dl$nameInfo$item, c('spont','control','evaluated','waiting')),
  responseNames=c("much more","somewhat more", 'equal',
                  "somewhat less", "much less"))
ggplot(df) +
  geom_line(aes(x=worthDiff,y=prob,color=response,linetype=response,
                group=responseSample), size=.2, alpha=.2) +
  xlab("difference in latent worths") + ylab("probability") +
  ylim(0,1) + facet_wrap(~item) +
    guides(color=guide_legend(override.aes=list(alpha = 1, size=1)))
```

These response curves are a function of the `thresholds`, `scale`,
and `alpha` parameters. A detailed description of the item
response model can be found in the man page for `responseCurve`.
A large `alpha` (>1) can mean that the item discriminates among
objects well. However, it can also mean that the model
predicts all responses will be _equal_. If most observed responses are
indeed _equal_ then this can result in good model fit, but we
can also infer that the item is useless.

```{r}
alpha <- summary(fit2, pars=c("alpha"), probs=c(.5))$summary
rownames(alpha) <- covItemNames
```

```{r, results='hide'}
print(alpha[alpha[,'sd']>.25,,drop=FALSE])
```
```{r, results='asis', echo=FALSE}
kable(alpha[alpha[,'sd']>.25,,drop=FALSE])
```

I already decided to exclude _waiting_ by inspection of
the correlation matrix.
Item _chatter_ is on the borderline, but we can retain it.

We will enter the `alpha` parameters into the factor model
as non-stochastic data.
Trying to estimate `alpha` in the factor model causes bias,
at least in the models that I have tried.
The factor model is prone to increase both `alpha` and the
magnitude of factor proportions at the expense of
`threshold` accuracy.
To treat `alpha` as non-stochastic reduces variability
in the factor model, but not by much.
Simulations indicate that the posterior distribution remains well calibrated.

I will fit model 'factor_ll' instead of 'factor' so I can use the
**loo** package to look for outliers.
We also need to take care that the data `pafp` matches, one-to-one,
the data seen by Stan so we can map back from the model
to the data. Hence, we update `pafp` using the usual the data cleaning sequence
of `filterGraph` and `normalizeData` and pass the result to `prepCleanData`.

Up until version 1.0.2, only a single factor model was available. As
of 1.0.3, the factor model supports an arbitrary number of factors and
arbitrary factor-to-item structure. In this example, we will
keep with the simplest factor model, a single factor that predicts all items.
We need to set a prior on the factor loadings. The prior is given as the standard
deviation of the normal prior for the logit transformed factor proportion.
Here's an example:

```{r}
factorProportion <- .3
factorScalePrior <- .9
dnorm(qlogis(0.5 + factorProportion/2), sd=factorScalePrior)
```

Enough data are available that we can use a factor scale prior of
1.0. If the model exhibited sampling difficulties (i.e. divergences)
then a smaller prior could help by constraining the factor proportion
closer to zero.

```{r factorData1, cache=TRUE}
pafp <- pafp[,c(paste0('pa',1:2),
             setdiff(covItemNames, c('spont','control','evaluated','waiting')))]
pafp <- normalizeData(filterGraph(pafp))
dl <- prepCleanData(pafp)
dl <- prepSingleFactorModel(dl, 1.0)
dl$scale <- result[match(dl$nameInfo$item, result$item), 'scale']
dl$alpha <- alpha[match(dl$nameInfo$item, rownames(alpha)), 'mean']
```
```{r, results='hide', include = FALSE}
rm(fit2)  # free up some memory
```

```{r factor, message=FALSE, results='hide', cache=TRUE}
fit3 <- pcStan("factor_ll", data=dl, include=FALSE, 
               pars=c('rawUnique', 'rawUniqueTheta', 'rawPerComponentVar',
	       'rawFactor', 'rawLoadings', 'rawFactorProp', 'rawNegateFactor', 'rawSeenFactor',
	       'unique', 'uniqueTheta'))
```

To check the fit diagnostics, we have to take care to
examine only the parameters of interest. The factor model outputs
many parameters that should not be interpreted. For example, we do not
care about `unique` or `uniqueTheta`, the unique scores.

```{r factorDiag1, cache=TRUE}
check_hmc_diagnostics(fit3)

interest <- c("threshold", "pathProp", "factor", "lp__", "log_lik")

allPars <- summary(fit3, pars=interest)$summary
print(min(allPars[,'n_eff']))
print(max(allPars[,'Rhat']))
```

Looks good! Let us see which data are the most unexpected by the model.
We create a `loo` object and inspect the summary output.

```{r factorLoo, cache=TRUE}
options(mc.cores=1)  # otherwise loo consumes too much RAM
kThreshold <- 0.4
l1 <- toLoo(fit3) 
print(l1)
```

```{r results='hide', echo=FALSE}
if (sum(pareto_k_values(l1)>.5) == 0) stop("No outliers?!")
```

The first thing to check is whether the minimum `n_eff` for the Pareto
$k$ is greater than 100 times the number of chains. It is. Next, we look at
`p_loo`, the effective number of parameters. In well behaving cases,
`p_loo` is less than the sample size and the number of parameters.
This also looks good. There are `r dl$NITEMS * dl$NPA` parameters
just for the unique scores.

**loo** identified `r sum(pareto_k_values(l1)>.5)` observations with a
Pareto $k>0.5$.
These observations can be regarded as outliers.
To find out the details, we pass the `loo` object to `outlierTable`.
We use a threshold of `r kThreshold` instead of 0.5 to ensure
that ample lines are shown.

```{r}
ot <- outlierTable(dl, l1, kThreshold)
```
```{r, results='hide'}
print(ot)
```
```{r, results='asis', echo=FALSE}
kable(ot, row.names=TRUE)
```

```{r}
xx <- which(ot[,'pa1'] == 'calisthenics' & ot[,'pa2'] == 'sex' & ot[,'item'] == 'body' & ot[,'pick'] == -2)
```

We will take a closer look at row `r xx`.
What does a `pick` of `r ot[xx,'pick']` mean? `Pick` numbers are converted
to response categories by adding the number of thresholds plus one.
There are two thresholds (_much_ and _somewhat_) so
3 + `r ot[xx,'pick']` = `r 3+ot[xx,'pick']`.
Looking back at our item response curve plot,
the legend gives the response category order from top (1) to bottom (5).
The first response category is _much more_.
Putting it all together we obtain an endorsement of
_`r ot[xx,'pa1']` involves much more of my body than `r ot[xx,'pa2']`_.
Specifically what about that assertion is unexpected?
We can examine how other participants have responded,

```{r}
pafp[pafp$pa1 == ot[xx,'pa1'] & pafp$pa2 == ot[xx,'pa2'],
     c('pa1','pa2', as.character(ot[xx,'item']))]
```

Hm, this is the only participant that compared `r ot[xx,'pa1']` and `r ot[xx,'pa2']`.
Let us look a little deeper to understand why this response
was unexpected.

```{r outlier, cache=TRUE}
loc <- sapply(ot[xx,c('pa1','pa2','item')], unfactor) 
exam <- summary(fit3, pars=paste0("theta[",loc[paste0('pa',1:2)],
                          ",", loc['item'],"]"))$summary
```
```{r, results='asis', echo=FALSE}
#exam <- data.frame(mean=c(0,0), '2.5%'=c(0,0), '97.5%'=c(0,0))
kable(exam)
```

Here we find that, based on comparisons with other activities,
`r ot[xx,'pa2']` was estimated to involve `r exam[2,'mean'] - exam[1,'mean']` units more of
the body than `r ot[xx,'pa1']`. However, note the wide posterior quantiles!
What are the sample sizes associated with these activities?

```{r}
sum(c(pafp$pa1 == ot[xx,'pa1'], pafp$pa2 == ot[xx,'pa1']))
sum(c(pafp$pa1 == ot[xx,'pa2'], pafp$pa2 == ot[xx,'pa2']))
```

Ah, no wonder, the body involvement 95% uncertainty interval for `r ot[xx,'pa2']`
is from `r exam[2,'2.5%']` to `r exam[2,'97.5%']`.
So there is practically no information.
We could continue our investigation by looking at which
responses justified these _body_ estimates.
However, let us move on and
plot the marginal posterior distributions of the factor proportions,

```{r pathProp, cache=TRUE}
pi <- parInterval(fit3, 'pathProp', dl$nameInfo$item, label='item')
pi <- pi[order(abs(pi$M)),]
pi$item <- factor(pi$item, levels=pi$item)

ggplot(pi) +
  geom_vline(xintercept=0, color="green") +
  geom_jitter(data=parDistributionFor(fit3, pi),
              aes(value, item), height = 0.35, alpha=.05) +
  geom_segment(aes(y=item, yend=item, x=L, xend=U),
               color="yellow", alpha=.5) +
  geom_point(aes(x=M, y=item), color="red", size=1) +
  theme(axis.title.y=element_blank())
```

Finally, we can plot the factor scores. By default, activities with small sample
sizes are retained by `filterGraph` if they connect other activities
because they contribute information to the model. However, when we
look at the per-activity factor scores, we can limit ourselves to
activities with a sample size of at least 11.

```{r activities, cache=TRUE}
pa11 <- levels(filterGraph(pafp, minDifferent=11L)$pa1) 
pick <- paste0('factor[',match(pa11, dl$nameInfo$pa),',1]')
pi <- parInterval(fit3, pick, pa11, label='activity')
pi <- pi[order(pi$M),]
pi$activity <- factor(pi$activity, levels=pi$activity)

ggplot(pi) +
  geom_vline(xintercept=0, color="green") +
  geom_jitter(data=parDistributionFor(fit3, pi, samples=250),
              aes(value, activity), height = 0.35, alpha=.05) +
  geom_segment(aes(y=activity, yend=activity, x=L, xend=U),
               color="yellow", alpha=.5) +
  geom_point(aes(x=M, y=activity), color="red", size=1) +
  theme(axis.title.y=element_blank()) 
```

And there you have it. If you have not done so already,
go find a dojo and commence study of martial arts!

# Technical notes

Given that my background is more in software than math,
I am not a fan of the greek letters used
with such enthusiasm by mathematicians.
When I name variables, I favor the expressive over the succinct.

If you read through the **Stan** models included with this package, you will find some
variables prefixed with `raw`. These are special variables
internal to the model. In particular, you should not try
to evaluate the $\widehat R$ or effective sample size
of `raw` parameters. These parameters are best excluded
from the sampling output.

## Unidim Adapt

| parameter | prior             | purpose                  |
|-----------|-------------------|--------------------------|
| threshold | normal(0,2)       | item response thresholds |
| theta     | normal(0,sigma)   | latent score             |
| sigma     | lognormal(1,1)    | latent score scale             |
| scale     | N/A               | latent score scaling constant  |

The 'unidim_adapt' model has a `varCorrection` constant
that is used to calibrate the `scale`. For all other models,
per-item `scale` must be passed in as data.
`scale` has no substantive interpretation, but it is used to
partition signal between object variance and item discrimination.
While object variance has no substantive interpretation,
item discrimination is interpretable.

## Unidim

| parameter | prior             | purpose                  |
|-----------|-------------------|--------------------------|
| threshold | normal(0,2)       | item response thresholds |
| alpha     | exponential(0.1)  | item discrimination    |
| theta     | normal(0,1)       | latent score    |

## Correlation

| parameter | prior             | purpose                              |
|-----------|-------------------|--------------------------------------|
| threshold | normal(0,2)       | item response thresholds             |
| alpha     | exponential(0.1)  | item discrimination    |
| thetaCor  | lkj(2)            | correlations between items           |
| theta     | _see below_       | latent score                |

Thresholds for all items are combined into a single vector.
The prior for `theta` is multivariate normal with correlations
`thetaCor` and scale 1.0.
Exclude `rawTheta` and `rawThetaCorChol` from sampling reports.

## Factor

| parameter      | prior               | purpose                              |
|----------------|---------------------|--------------------------------------|
| threshold      | normal(0,2)         | item response thresholds             |
| unique         | normal(0,2)         | scale of unique scores               |
| uniqueTheta    | normal(0,1)         | unique scores                        |
| factorLoadings | normal(0,2)         | signed scale of factor scores        |
| factor         | normal(0,1)         | factor scores                        |
| pathProp       | _see below_         | signed factor variance proportion    |
| sigma          | N/A                 | relative item scale                  |

Thresholds for all items are combined into a single vector.
`pathProp` is computed based on Equation 3 of Gelman et al. (in press).
The prior on `pathProp` is `normal(logit(0.5 + pathProp/2), pathScalePrior)`
where `pathScalePrior` is provided as data (see `prepSingleFactorModel()`).
The idea of putting a prior on `pathProp` was inspired by Gelman (2019, Aug 23).
The `unique` and `factorLoadings` parameters are only relatively interpretable because
the scale is arbitrary.
The most interpretable parameter is `pathProp`.
`pathProp` is a signed proportion bounded between -1 and 1.
Exclude `rawUnique`, `rawUniqueTheta`, `rawFactor`, `rawLoadings`, `rawPerComponentVar`,
`rawPathProp`, `rawNegateFactor`, and `rawSeenFactor` from sampling.
You may also like to exclude `unique` and `uniqueTheta` unless you have
some special interest in the unique scores.

# References

Gelman, A. (2019, Aug 23). Yes, you can include prior information on quantities of interest, not just on parameters in your model [Blog post]. Retrieved from [https://statmodeling.stat.columbia.edu/2019/08/23/yes-you-can-include-prior-information-on-quantities-of-interest-not-just-on-parameters-in-your-model/](https://statmodeling.stat.columbia.edu/2019/08/23/yes-you-can-include-prior-information-on-quantities-of-interest-not-just-on-parameters-in-your-model/).

Gelman, A., Goodrich, B., Gabry, J., & Vehtari, A. (in press). R-squared for Bayesian regression models. _The American Statistician._ DOI: 10.1080/00031305.2018.1549100

Silver, D., Hubert, T., Schrittwieser, J., Antonoglou, I., Lai, M.,
Guez, A., ... & Lillicrap, T. (2018). A general reinforcement
learning algorithm that masters chess, shogi, and Go through
self-play. Science, 362(6419), 1140-1144.

Vehtari, A., Gelman, A., Simpson, D., Carpenter, B., & Bürkner, P. C. (2019). Rank-normalization, folding, and localization: An improved $\widehat R$ for assessing convergence of MCMC. _arXiv preprint_ arXiv:1903.08008.

# R Session Info

```{r}
sessionInfo()
```
